---
title: "How to make a plot by ggplot2"
author: "Wenyu Zhu"
date: "29 Oct, 2019"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

# Get started
## Installation
We assume you have already installed R (https://www.r-project.org) (R Core Team 2019) and the RStudio IDE (https://www.rstudio.com). Next we need to install the tidyverse package and mdsr package in R

```{r,eval=FALSE}
install.packages('ggplot2')
install.packages('mdsr')
```

```{r,eval=FALSE}
library(ggplot2)
library(mdsr)
```

## A quick example
```{r}
curve(dnorm, from=-4, to=4, ylab="PDF", main="PDF of a standard normal", col="brown", lwd=4)
```

This is the simplest chart without importing data.

# Make a plot

## ggplot2

In ggplot2

- the ggplot() command creates a plot, with the argument data telling us which data to use

- the aes() command within ggplot() indicates the desired x and y variables

- plots are built with elements

- the element geom_point() indicates that we are using a scatterplot

We will be using the data set CIACountries from the mdsr package. The CIACountries data table contains the following seven variables for 236 countries:

- population (pop)

- area (area)

- gross domestic product (gdp)

- percentage of GDP spent on education (educ)

- length of roadways per unit area (roadways)

- internet use as a fraction of the population (net_users)

- number of barrels of oil produced per day (oil_prod)

```{r,include=FALSE}
library(mdsr)
```
```{r}
head(CIACountries)
```

Make a scatterplot of Gross Domestic Product and Education.
```{r}
g <- ggplot(data = CIACountries, aes(y=gdp, x = educ)) +
  geom_point(na.rm=TRUE)+
theme(axis.text=element_text(size=16),
axis.title=element_text(size=20),
plot.title=element_text(size = 24)) +
ylab('Gross Domestic Product') +
xlab('Percentage of Gross Domestic \n Product Spent on Education') +
ggtitle('Gross Domestic Product and Education')
g
```

Scatterplot in which net_users is mapped to color.
```{r}
g + geom_point(aes(color = net_users), size = 3,na.rm=TRUE) +
guides(color=guide_legend(title="Internet Use"))
```

Scatterplot using both location and label as aesthetics.
```{r,warning=FALSE}
g + geom_text(aes(label = country, color = net_users), size = 3) +
guides(color=guide_legend(title="Internet Use"))
```

Scatterplot in which net_users is mapped to color and educ mapped to size.

```{r,warning=FALSE}
g + geom_point(aes(color = net_users, size = roadways)) +
guides(color=guide_legend(title="Internet Use"),
size=guide_legend(title='Roadway'))
```

## Scale

From the graphs that we made it is difficult to discern differences in GDP because there is a right-skew in GDP and we made a choice to use a linear scale on our plots.

- A logarithmic scale would make the plot much more readable.

- We will use the coord_trans() function to put our data on a logarithmic scale.

Scatterplot using the logarithmic transform of GDP that helps to mitigate visual clustering caused by the right-skewed distribution of GDP among cities.

```{r,warning=FALSE}
g + geom_point(aes(color = net_users, size = roadways)) +
guides(color=guide_legend(title="Internet Use"),
size=guide_legend(title='Roadway')) +
coord_trans(y = "log10")
```

As an alternative to coord_trans() we can use the scale_y_continuous() function. The only difference is where the tick marks on the axis are drawn.

The functions scale_x_continuous(), scale_x_discrete(), and scale_color() all perform analogous operations on different aesthetics.

Scatterplot using the logarithmic transform of GDP that helps to mitigate visual clustering caused by the right-skewed distribution of GDP among cities.

```{r,warning=FALSE}
g + geom_point(aes(color = net_users, size = roadways)) +
guides(color=guide_legend(title="Internet Use"),
size=guide_legend(title='Roadway')) +
scale_y_continuous(name = 'Gross Domestic Product', trans = 'log10')
```

## Facets

Using multiple aesthetics such as shape, color, and size to display multiple variables can produce a really hard to read graph!

Facets are multiple side by side graphs used to display levels of categorical variables which can help.

Scatterplot using facets for different ranges of Internet connectivity.

```{r,warning=FALSE}
g + geom_point(alpha = 0.9, aes(size = roadways)) +
coord_trans(y = 'log10') +
facet_wrap(~net_users, nrow = 1) +
theme(legend.position = 'top') +
theme(axis.text.x =element_text(size=10))
```

## Layers

We can plot data from more than one data table together.

We will now look at the MedicareProviders data set for an example. This data set tells how much different medical procedures cost in each state. The data set contains the following variables:

- medical procedure code (drg)

- state (stateProvider)

- number of charges with corresponding code (num_charges)

- mean charge for corresponding code (mean_charge)

```{r}
head(MedicareCharges)
```

Let’s create a new data table with only the data from New Jersey. Bar graph of average charges for medical procedures in New Jersey

```{r}
ChargesNJ <- MedicareCharges %>% filter(stateProvider == 'NJ')
p <- ggplot(data = ChargesNJ,
aes(x = reorder(drg, mean_charge),
y = mean_charge)) +
geom_bar(fill = 'gray', stat = 'identity') +
ggtitle('Average Medical Costs by Procedure in New Jersey') +
ylab('Statewide Average Charges ($)') +
xlab('Medical Procedure (DRG)') +
theme(axis.text.y =element_text(size=16),
axis.title=element_text(size=20),
plot.title = element_text(size = 24),
axis.text.x = element_text(angle = 90, hjust = 1))
p
```

Bar graph adding a second layer to provide comparison of New Jersey to other states. Each dot represents one state while the bars represent Ne Jersey.

```{r}
p + geom_point(data = MedicareCharges, size = 1, alpha = 0.3)
```

# Several kinds of plot

## Histogram

We will use the math SAT scores (math) from each state in the SAT_2010 data set.

Histogram showing the distribution of Math SAT scores by state.

```{r}
head(SAT_2010)
```

```{r}
g <- ggplot(data = SAT_2010, aes(x = math)) +
theme(axis.text=element_text(size=16),
axis.title=element_text(size=20),
plot.title=element_text(size = 22)) +
xlab('Math SAT Score') +
ggtitle('State Average Math Scores on the 2010 SAT')
g + geom_histogram(binwidth = 10)
```

## Density

Density plot showing the distribution of the Math SAT scores by state.

```{r}
g + geom_density(adjust = 0.3)
```

## Bar graphs

A bar plot showing the distribution of MATH SAT scores for a selection of states.

```{r}
ggplot(data = head(SAT_2010, 10), aes(x = reorder(state, math), y = math)) +
geom_bar(stat = 'identity') +
theme(axis.text=element_text(size=16),
axis.title=element_text(size=20),
plot.title=element_text(size = 22)) +
xlab('Math SAT Score') +
ggtitle('State Average Math Scores on the 2010 SAT for Top 10 States')
```

## Scatterplot

Recall our SAT_2010 example:

```{r}
head(SAT_2010)
```

A scatterplot of the SAT math scores and expenditure by pupil by states.

```{r}
g <- ggplot(data = SAT_2010, aes (x = expenditure, y = math)) +
geom_point()
g
```

A scatterplot of the SAT math scores and expenditure by pupil by states with a smooth trend line.

```{r}
g <- g + geom_smooth(method = 'lm', se = 0) +
xlab("Average expenditure per student ($1000)") +
ylab("Average score on math SAT") +
ggtitle('Average Expediture versus Average SAT Score by State') +
theme(axis.text=element_text(size=14),
axis.title=element_text(size=20),
plot.title=element_text(size = 22))
g
```

We will now place states into bins (e.g., high, medium, low) by the percentage of students taking the SAT.

Scatterplot using the color aesthetic to separate the relationship between numeric variables by a third categorical variable.

```{r}
SAT_2010 <- SAT_2010 %>%
mutate(SAT_rate = cut(sat_pct, breaks = c(0, 30, 60, 100),
labels = c('low', 'medium', 'high')))
g <- g %+% SAT_2010
g + aes(color = SAT_rate)
```

Scatterplot using facet_wrap() to separate the relationship between two numeric variables by a third categorical variable.

```{r}
g + facet_wrap(~SAT_rate)
```

## Boxplot

Boxplots can be useful for displaying a numerical response variable against a categorical explanatory variable. It is a graphic description of the five-number summary (minimum, Q1, median, Q3, and maximum).

We will look at the length of children’s feet (in cm) by sex.

```{r}
library(mosaic)
favstats(length ~ sex, data = KidsFeet)
```

A box-and-whisker plot (boxplot) showing the distribution of foot length by gender for 39 children.

```{r}
ggplot(data = KidsFeet, aes(x = sex, y = length)) +
geom_boxplot()
```

This boxplot is not very well. Based on the knowledge above, how can we improve the plot?

# A cool plot

Use the make_babynames_dist() function in the mdsr package to recreate the “Deadest Names” graphic from FiveThirtyEight.

Follow below steps:

• Using what we have learned about the ggplot2.

- (a) make a dataframe with only females 

- (b) calculate the total number of each name from 1900 to the present 

- (c) only take names having over 100,000 people having that name from 1900 to present (i.e. common names).

- (d) calculate the percent of people with that name who are dead. 

- (e) select the top 10 names.

• Do the same for males.

• Bind the two dataframes you created together.

• Use geom_bar and coord_ﬂip to get started on the plot!

```{r,warning=FALSE}
library(babynames)
library(mdsr)

BabynamesDist <- make_babynames_dist()
deadest_female <- BabynamesDist %>%
filter(sex == 'F') %>% ##filter for only females
group_by(name) %>% ## group by names
mutate(N = sum(count_thousands * 1000 )) %>% ##calculate the total number with each name from 1900 to present
filter(N >= 100000) %>% ##filter so we only have common names
summarise(percent_dead_today = 1 - sum(est_alive_today) / sum(count_thousands * 1000 )) %>% ## calc percent dead
ungroup() %>% ##ungroup
arrange(desc(percent_dead_today)) %>% ##arrange by percent dead in descending order
mutate(sex = 'F') %>% ## add sex back in
head(10) ##select the top 10

deadest_male <- BabynamesDist %>%
filter(sex == 'M') %>% ##filter for only males
group_by(name) %>% ## group by names
mutate(N = sum(count_thousands * 1000 )) %>% ##calculate the total number with each name from 1900 to present
filter(N >= 100000) %>% ##filter so we only have common names
summarise(percent_dead_today = 1 - sum(est_alive_today) / sum(count_thousands * 1000 )) %>% ## calc percent dead
ungroup() %>% ##ungroup
arrange(desc(percent_dead_today)) %>% ##arrange by percent dead in descending order
mutate(sex = 'M') %>% ## add sex back in
head(10) ##select the top 10

deadest <- rbind(deadest_female,
deadest_male) ##bind the two dataframes togather
deadest <- deadest %>%
mutate(label = paste0(round(percent_dead_today * 100, 1), '%')) ##add a variable label which will be the percent dead
```

```{r}
ggplot(deadest, aes(x = reorder(name, percent_dead_today), ##reorder name by the percent dead today
y = percent_dead_today,
fill = sex, ##color the plot by sex
label = label)) + ##add label for the percent dead
geom_bar(stat = 'identity',) + ## make the barplot
coord_flip() + ## flip the coordinates
scale_fill_manual(values=c('#e1ad01', '#1C86EE')) + ## change the colors
geom_text( size = 3, position = position_stack(vjust = 1.05)) + ## add the label for the percent dead today
theme(legend.position="none", ##remove the legend
panel.grid.major = element_blank(), panel.grid.minor = element_blank(), ##remove the background grid
axis.text.x=element_blank(), ## remove the x-axis text
axis.ticks.x=element_blank()) + ## remove the x-axis tick marks
labs(x= '', ##no label for x-axis
y = '', ##no label for y-axis
title = 'Deadest Names', ##add title
subtitle = 'Estimated percentage of Americans with a given name born since \n1900 who were dead as of Jan. 1, 2014.')
```
